---
title: "Data Cleaning Checklist"
author: "Jason Entingh"
date: "`r Sys.Date()`"
output: html_document
---

# setup

## markdown setup

```{r setup_markdown, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## library

```{r setup_library}
library(tidyverse)
library(readxl)
library(writexl)
library(nycflights13)
library(janitor)
```

# cleaning checklist

## check data types

```{r cleaning_datatypes}

planes |> 
  glimpse()

```

## duplicates

```{r cleaning_duplicates}

# get summary of duplicates
planes |> 
  count(tailnum) |> 
  filter(n > 1)

# find duplicates based on variables
planes |> 
  group_by(tailnum) |> 
  filter(n() > 1)

# remove duplicates
planes |> 
  distinct(tailnum, .keep_all = TRUE)

```

## missing values

### explicit

Explicit missing values are those noted by the presence of an NA value. 

An explicit missing value is the presence of an absence.

```{r cleaning_missing_explicit}

# simple summary
planes |> 
  summary()

# summarize observations and missing counts for each variable
planes |> 
  summarise(across(everything(),
                   list(
                     n_obs = ~ length(.),
                     n_missing = ~ sum(is.na(.))
                   ))) |> 
  pivot_longer(everything(),
               names_to = c("variable", "stat"),
               names_pattern = "(.*)_(n_obs|n_missing)",
               values_to = "value") |> 
  pivot_wider(names_from = stat, values_from = value)

# if working with years, can then check if any particular year has a higher proportion of missing values for a particular variable
planes |> 
  group_by(year) |> 
  summarize(
    n = n(),
    sum_missing = sum(is.na(speed))
  )

```

### implicit

Implicit missing values are those where an observation doesn't show up in the dataset.

An implicit missing value is the absence of a presence.

There are a couple of ways to handle this, via transforming implicit missing values to explicit missing values.
* Pivoting data wider  
* Using the `complete()` function  

```{r cleaning_missing_implicit}

# create sample dataframe
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

# pivot wider
stocks |>
  pivot_wider(
    names_from = qtr, 
    values_from = price
  )

# complete function
stocks |>
  complete(year, qtr)

```

### missing data diagnostic

The following code is a diagnostic built to detect patterns of missing data by year for each entity, such that each entity ends up in a matrix of the following:  
* all NA: entity is missing all years of data  
* complete: entity has all years of data  
* leading only: entity is missing data only at the beginning (e.g. no data from 2019-2021, then data in 2022, 2023, and 2024) - candidate for backward factor imputation
* internal only: entity is missing data 'holes' - candidate for interpolation/linear imputation  
* trailing only: entity is missing data only at the end (e.g. only in 2024, or only in 2023-2024, or only in 2022-2024...) - candidate for forward factor imputation  
* mixed: entity is missing data 'holes', and/or at the end, and/or at the beginning (e.g. missing data in 2022 and 2024, or missing data in 2021, 2023-2024)  

Data needs to be complete (no implicit missing values) and in a long format.

```{r cleaning_missing_diagnostic}

# get sample dataset
library(plm)
data("Grunfeld")

# insert NAs
set.seed(3)
grunfeld_missing <- Grunfeld |> 
  filter(year >= 1940 & year <= 1945)

grunfeld_missing$value[sample(nrow(grunfeld_missing), 10)] <- NA

summary(grunfeld_missing)

# we can get a general idea of how much data is missing in each year by doing a simple summarize command
grunfeld_missing |> 
  group_by(year) |> 
  summarize(
    n = n(),
    sum_missing = sum(is.na(value))
  )

grunfeld_missing_diagnostic <- grunfeld_missing |> 
  arrange(firm, year) |> 
  group_by(firm) |> 
  mutate(
    na_flag = is.na(value),
    # identify leading NAs
    leading_na = {
      na_vec <- is.na(value)
      # do not mark as leading if all values are missing
      if (all(na_vec)) {
        rep(FALSE, length(na_vec))
      } else {
      cumall(na_vec)
      }
    },
    # identify trailing NA block
    trailing_na = {
      na_vec <- is.na(value)
      # do not mark as trailing if all values are missing
      if (all(na_vec)) {
        rep(FALSE, length(na_vec))
        # if all values not missing, go for it
        # reverse logical flag so trailing years are first
        # cumall() reads the na logic check, returning TRUE for TRUE, until it reads a FALSE value - after which it returns all FALSE. 
        # reverse again.
        # this lets us detect and mark NAs that are the trailing NAs of an entity 
      } else {
        rev(cumall(rev(na_vec)))
      }
    },
    # internal NAs
    internal_na = {
      na_vec <- is.na(value)
      # has non-NA before
      has_before <- cumany(!na_vec)
      # has a non-NA after
      has_after <- rev(cumany(rev(!na_vec)))
      # internal NA definition
      na_vec & !trailing_na & has_before & has_after
    },
  ) |> 
  ungroup() |> 
  group_by(firm) |> 
  mutate(
    pattern = case_when(
      # all NA - all values missing
      all(is.na(value)) ~ "all NA",
      # complete - all values nonmissing
      all(!is.na(value)) ~ "complete",
      # leading NAs - has leading NAs, and nonmissing values after.
      any(leading_na) & !any(internal_na) & !any(trailing_na) ~ "leading_only",
      # trailing only - has nonmissing from start, then all NAs after
      any(trailing_na) & !any(internal_na) & !any(leading_na) ~ "trailing_only",
      # internal only - has complete data from start, then has a data hole (of any size, as long as there's data after), and no trailing NAs.
      !any(trailing_na) & any(internal_na) & !any(leading_na) ~ "internal_only",
      # has a mix of trailing NAs, internal NAs, and leading NAs.
      any(trailing_na) & any(internal_na) | any(trailing_na) & any(leading_na) | any(leading_na) & any(internal_na) ~ "mixed",
      # the rest are EINs that closed at some point.
      !any(leading_na) & !any(internal_na) & !any(trailing_na) & any(na_flag) ~ "closed"
    )
  ) |> 
  ungroup() |> 
  select(firm, year, value, na_flag:pattern)

grunfeld_missing_diagnostic |> 
  tabyl(pattern)


```

## logical consistency

Be sure to check that related variables are consistent. Examples include:
* date relationships (e.g. end_date > start_date)  
* age vs birth year  
* mutually exclusive fields  
* conditional missingness  
* range dependency on category  
* total vs components  
* binary flags vs counts  

### validation function

```{r cleaning_consistency_validationfunction}

# this validation function was built using ChatGPT.
## validation function
validate_data <- function(df) {
  df %>%
    mutate(
      # --- Date checks ---
      check_date_order = if ("start_date" %in% names(.) & "end_date" %in% names(.)) {
        end_date < start_date
      } else FALSE,

      # --- Age consistency ---
      check_age = if ("age" %in% names(.) & "birth_year" %in% names(.)) {
        abs(age - (as.numeric(format(Sys.Date(), "%Y")) - birth_year)) > 1
      } else FALSE,

      # --- Conditional missingness ---
      check_car = if ("has_car" %in% names(.) & "car_model" %in% names(.)) {
        has_car == TRUE & is.na(car_model)
      } else FALSE,

      # --- Logical consistency ---
      check_children = if ("has_children" %in% names(.) & "num_children" %in% names(.)) {
        has_children == FALSE & num_children > 0
      } else FALSE,

      # --- Totals check ---
      check_total = if (all(c("total", "part_a", "part_b") %in% names(.))) {
        total != part_a + part_b
      } else FALSE
    )
}

## summarize issues
summarize_issues <- function(df_checked) {
  df_checked %>%
    summarise(across(starts_with("check_"), ~ sum(.x, na.rm = TRUE)))
}

## view problematic rows
get_problem_rows <- function(df_checked) {
  df_checked %>%
    filter(if_any(starts_with("check_"), ~ .x))
}

## apply
df_checked <- validate_data(df)

# See counts of issues
summarize_issues(df_checked)

# Inspect bad rows
bad_rows <- get_problem_rows(df_checked)



```

### R validate package

```{r cleaning_consistency_validatepackage}

# The Data Validation Cookbook: https://data-cleaning.github.io/validate/
# github repo: https://github.com/data-cleaning/validate

library(validate)

# set rules
rules <- validator(speed >= 200, 
                   seats >= 200)

# compare dataset to rules
out <- confront(planes, rules)

# see results
summary(out)

# visualize results
plot(out)


```

## clean character values

### check consistency of character values

```{r cleaning_chr_consistency}
# single variable
planes |> 
  tabyl(engine)

```

## clean numeric values

### check outliers and distribution

```{r cleaning_num_outliers}

# simple summary
planes |> 
  summary()

# box plots
## boxplot function
planes |> 
  select(where(is.numeric)) |> 
  map(~ boxplot(.x, main = deparse(substitute(.x))))

## ggplot
planes |> 
  pivot_longer(cols = where(is.numeric)) |> 
  ggplot(aes(y = value)) +
  geom_boxplot() +
  facet_wrap(~ name, scales = "free")

# histogram
## basic histogram
planes |> 
  select(where(is.numeric)) |> 
  imap(~ hist(.x, main = .y, xlab = .y))

## ggplot
planes |> 
  pivot_longer(cols = where(is.numeric)) |> 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ name, scales = "free")

```
