---
title: "Data Cleaning Checklist"
author: "Jason Entingh"
date: "`r Sys.Date()`"
output: html_document
---

# setup

## markdown setup

```{r setup_markdown, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## library

```{r setup_library}
library(tidyverse)
library(readxl)
library(writexl)
library(nycflights13)
library(janitor)
```

# cleaning checklist

## check data types

```{r cleaning_datatypes}

planes |> 
  glimpse()

```

## duplicates

```{r cleaning_duplicates}

# get summary of duplicates
planes |> 
  count(tailnum) |> 
  filter(n > 1)

# find duplicates based on variables
planes |> 
  group_by(tailnum) |> 
  filter(n() > 1)

# remove duplicates
planes |> 
  distinct(tailnum, .keep_all = TRUE)

```

## missing values

### explicit

Explicit missing values are those noted by the presence of an NA value. 

An explicit missing value is the presence of an absence.

```{r cleaning_missing_explicit}

# simple summary
planes |> 
  summary()

```

### implicit

Implicit missing values are those where an observation doesn't show up in the dataset.

An implicit missing value is the absence of a presence.

There are a couple of ways to handle this, via transforming implicit missing values to explicit missing values.
* Pivoting data wider  
* Using the `complete()` function  

```{r cleaning_missing_implicit}

# create sample dataframe
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)

# pivot wider
stocks |>
  pivot_wider(
    names_from = qtr, 
    values_from = price
  )

# complete function
stocks |>
  complete(year, qtr)

```

### missing data diagnostic

Find a long dataset that I can use the missing data panel check on (what I made for DAF)

```{r cleaning_missing_diagnostic}




```


## logical consistency

Be sure to check that related variables are consistent. Examples include:
* date relationships (e.g. end_date > start_date)  
* age vs birth year  
* mutually exclusive fields  
* conditional missingness  
* range dependency on category  
* total vs components  
* binary flags vs counts  

### validation function

```{r cleaning_consistency_validationfunction}

# this validation function was built using ChatGPT.
## validation function
validate_data <- function(df) {
  df %>%
    mutate(
      # --- Date checks ---
      check_date_order = if ("start_date" %in% names(.) & "end_date" %in% names(.)) {
        end_date < start_date
      } else FALSE,

      # --- Age consistency ---
      check_age = if ("age" %in% names(.) & "birth_year" %in% names(.)) {
        abs(age - (as.numeric(format(Sys.Date(), "%Y")) - birth_year)) > 1
      } else FALSE,

      # --- Conditional missingness ---
      check_car = if ("has_car" %in% names(.) & "car_model" %in% names(.)) {
        has_car == TRUE & is.na(car_model)
      } else FALSE,

      # --- Logical consistency ---
      check_children = if ("has_children" %in% names(.) & "num_children" %in% names(.)) {
        has_children == FALSE & num_children > 0
      } else FALSE,

      # --- Totals check ---
      check_total = if (all(c("total", "part_a", "part_b") %in% names(.))) {
        total != part_a + part_b
      } else FALSE
    )
}

## summarize issues
summarize_issues <- function(df_checked) {
  df_checked %>%
    summarise(across(starts_with("check_"), ~ sum(.x, na.rm = TRUE)))
}

## view problematic rows
get_problem_rows <- function(df_checked) {
  df_checked %>%
    filter(if_any(starts_with("check_"), ~ .x))
}

## apply
df_checked <- validate_data(df)

# See counts of issues
summarize_issues(df_checked)

# Inspect bad rows
bad_rows <- get_problem_rows(df_checked)



```

### R validate package

```{r cleaning_consistency_validatepackage}

# The Data Validation Cookbook: https://data-cleaning.github.io/validate/
# github repo: https://github.com/data-cleaning/validate

library(validate)

# set rules
rules <- validator(speed >= 200, 
                   seats >= 200)

# compare dataset to rules
out <- confront(planes, rules)

# see results
summary(out)

# visualize results
plot(out)


```

## clean character values

### check consistency of character values

```{r cleaning_chr_consistency}
# single variable
planes |> 
  tabyl(engine)

```

## clean numeric values

### check outliers and distribution

```{r cleaning_num_outliers}

# simple summary
planes |> 
  summary()

# box plots
## boxplot function
planes |> 
  select(where(is.numeric)) |> 
  map(~ boxplot(.x, main = deparse(substitute(.x))))

## ggplot
planes |> 
  pivot_longer(cols = where(is.numeric)) |> 
  ggplot(aes(y = value)) +
  geom_boxplot() +
  facet_wrap(~ name, scales = "free")

# histogram
## basic histogram
planes |> 
  select(where(is.numeric)) |> 
  imap(~ hist(.x, main = .y, xlab = .y))

## ggplot
planes |> 
  pivot_longer(cols = where(is.numeric)) |> 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ name, scales = "free")

```
